Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-07-14 ltx/data_paper/DataPaper.tex
edits to result section
edits to discussion

$\Xi_0, \Xi_1, \Xi_2$&\eqref{eq:Xi0Gibson}, \eqref{eq:Xi1Gibson}, \eqref{eq:Xi2Gibson}&$O(Tn_x^2)$\\
$\Xi_3, \Xi_5$&\eqref{eq:Xi3Ap},\eqref{eq:Xi5Ap} &$O(Tn_x^3)+O(Tn_x^2n_r)+O(Tn_xn_{\theta}n_r)+O(Tn_{\theta}n_r^2)$ \\
$\Xi_4$&\eqref{eq:Xi4Ap}&$O(Tn_x^4n_{\theta})+O(Tn_x^3n_{\theta}^2)+O(Tn_x^3n_{\theta}n_r^2)$\\

@@

The effect of the observation noise variance on the connectivity kernel support is very much like to that of disturbance covariance function where an initial guess can be made in a similar manner. To study how synaptic parameter, $\xi=1-\frac{ T_s}{\tau}$, influence the resultant support of the connectivity kernel different values of synaptic time constatnt, $\tau$, were considered in a physiologically plausible range while noting that its value should be also higher than the sampling time. In fact synaptic parameter in \eqref{eq:EM-Fourier_TF_of_Kernel}  acts like a scaling factor, dilating or shrinking the estimated support, as the subtracting a constant in a frequency domain only affects the value at the origin in the spatial domain. The result of the correlation analysis for the connectivity kernel is shown in Figure~\ref{fig:KernelWidthEstimationExperiment} where each subplot is the normalised  estimate for varying synaptic time constants while the observation noise was kept constant. Note that for clarification the amplitude at the origin was negated when the chosen values of $\tau$ and $\sigma^2_{\varepsilon}$ resulted in the negative values at the origin.

@@

The estimated supports were then employed in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated kernel support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.
\section{Discussion}

\begin{itemize}
	\item comment on how the e step is UKF and M-step is kinda like the EKF, and we dont use EKF in E-step since UKF gives better estimates and the linearisation in M-step would be more accurate. 
	\item comment on EM has been used with EKS by Ghahramani which has poorer convergance property than using EM with URTSS
	\item itterative EM and correlation analysis (maybe)
	\item{comment on EM might convrge to suboptimal solution,simulated
annealing can be used to tackle such situations but difficult to implement (Mc Lachlan)}
	\item{since the non-linearity is a sigmoid function first-order Taylor provide a good approximation however in general settings with more complicated nonlinear behaviours higher order terms might need to be included, this however introduce intractable computational complexity.}
\end{itemize}

>>>>

$\mathbf q(\mathbf{x}_t)$&\eqref{}&$O(n_xn_{\theta}n_r)$\\
$\boldsymbol\Xi_0, \boldsymbol\Xi_1, \boldsymbol\Xi_2$&\eqref{eq:Xi0Gibson}, \eqref{eq:Xi1Gibson}, \eqref{eq:Xi2Gibson}&$O(Tn_x^2)$\\
$\boldsymbol\Xi_3, \boldsymbol\Xi_5$&\eqref{eq:Xi3Ap},\eqref{eq:Xi5Ap} &$O(Tn_x^3)+O(Tn_x^2n_r)+O(Tn_xn_{\theta}n_r)$ \\
$\boldsymbol\Xi_4$&\eqref{eq:Xi4Ap}&$O(Tn_x^4n_{\theta})+O(Tn_x^3n_{\theta}^2)+O(Tn_x^3n_{\theta}n_r)$\\

@@

The effect of the observation noise variance on the connectivity kernel support is very much like to that of disturbance covariance function where an initial guess can be made in a similar manner. To study how synaptic parameter, $\xi=1-\frac{ T_s}{\tau}$, influence the resultant support of the connectivity kernel different values of synaptic time constatnt, $\tau$, were considered in a physiologically plausible range while noting that its value should be also higher than the sampling time. In fact synaptic parameter in \eqref{eq:EM-Fourier_TF_of_Kernel}  acts like a scaling factor, dilating or shrinking the estimated support, as the subtracting a constant in a frequency domain only affects the value at the origin in the spatial domain. The result of the correlation analysis for the connectivity kernel is shown in Figure~\ref{fig:KernelWidthEstimationExperiment} where each subplot is the normalised  estimate for varying synaptic time constants while the observation noise was kept constant. Note that for clarification the amplitude at the origin was negated when the chosen values of $\tau$ and $\sigma^2_{\varepsilon}$ resulted in the negative values at the origin. The result suggests that choosing the range of approximately $\left[-5,5 \right] $ is reasonable for the spatial extent of the connectivity kernel.

@@

The estimated supports were then employed in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated kernel support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.
\section{Discussion}
The correlation analysis specifies approximate supports for both disturbance covariance function and connectivity kernel. Since the analysis assumes the sigmoidal firing rate to be in the active region, the amplitudes for the kernel and the disturbance cannot be inferred this way. Although the unknown parameters $\xi$ and $\sigma_{\varepsilon}^2$ can be chosen as described in Correlation analysis section, their values influence the exact shape of the kernel significantly, this step might be considered as an initialisation for the EM algorithm. Knowing the spatial extent of the kernel greatly increases the speed of the algorithm as the number of basis functions to decompose the kernel reduces significanly compared to the case in which the basis funtions are laid out across the entire spatial domain. This also improves the uncertainty of the parameter estimates as the number of parameters to be estimated reduces. On the other hand, the approximated support of the disturbance covariance function facilitates the implementation of the EM algorithm as the disturbance covariance matrix can be factorised into a constant matrix which only depends on the inferred support and an unknown scaling parameter which needs to be estimated.

The additional Taylor expansion step inroduced in the E-step resembles the extended Kalman filtering (EKF) \cite{Jazwinski1970,Chui2009} in which a stationary nonlinear dynamic system is approximated by a linear non-stationary system using the first order Taylor approximation of the state and observation equations about the state estimates and then applying regular Kalman filter to the linear system. This may suggest that the extended Kalman smoother is a natural approach to infer the sequence of states in the presented framework. However, extended Kalman smoother (EKS) requires relinearising about the backwards estimates which in turn necessitates the nonlinear mapping to be uniquely invertible. Although this is indeed the case in the neural field equation employed in this work it will impose more computational complexity in the state estimation procedure. In addition, URTSS provides better approximation compared to the EKS and therefore the approximated lower bound will be more accurate \cite{Haykin2001}.

\begin{itemize}
	\item comment on EM has been used with EKS by Ghahramani which has poorer convergance property than using EM with URTSS
	\item itterative EM and correlation analysis (maybe)
	\item{comment on EM might convrge to suboptimal solution,simulated
annealing can be used to tackle such situations but difficult to implement (Mc Lachlan)}
	\item{since the non-linearity is a sigmoid function first-order Taylor provide a good approximation however in general settings with more complicated nonlinear behaviours higher order terms might need to be included, this however introduce intractable computational complexity.}
\end{itemize}
